### 2022-02-28 doc/src/AdvPLIC.tex
Adopt "APLIC" as common name for Advanced PLIC

%=======================================================================
\chapter{Advanced Platform-Level Interrupt Controller (PLIC)}
\label{ch:AdvPLIC}
\chaptermark{Advanced PLIC}

@@

In a machine without IMSICs, every {\RISCV} hart accepts interrupts
from exactly one PLIC that is the \emph{external interrupt controller}
for that hart.
A hart's external interrupt controller (the PLIC) signals interrupts
to the hart through a dedicated connection, usually a wire, for each
privilege level that the hart may receive interrupts.
(Recall Figure~\ref{fig:intrsWithoutIMSICs} on
page~\pageref{fig:intrsWithoutIMSICs}.)
A system without IMSICs will typically have only one PLIC, serving as
the external interrupt controller for all {\RISCV} harts.

\begin{commentary}
Because every {\RISCV} hart without an IMSIC has exactly one PLIC
as its external interrupt controller, a system with multiple PLICs
must partition the harts into disjoint subsets, making each PLIC the
external interrupt controller for a separate subset of the harts.
While not prohibited, this arrangement is likely to be less efficient
than having all harts share a single PLIC.
\end{commentary}

{\RISCV} harts that employ IMSICs as their external interrupt
controllers can receive external interrupts only in the form of MSIs.
In that case, the role of a PLIC is to convert wired interrupts into
MSIs for harts.
(Recall Figure~\ref{fig:intrsWithIMSICs} on
page~\pageref{fig:intrsWithIMSICs}.)
The PLIC is said to \emph{forward} incoming wire-signaled interrupts to
harts by sending MSIs to the harts.

When harts have IMSICs to support MSIs, a system may easily contain
multiple PLICs for converting wired interrupts into MSIs, with each
PLIC forwarding interrupts from a different subset of devices.
Multiple PLICs are presumably more likely to arise when groups of
devices are physically distant from one another, perhaps even on
separate chips (including chiplets in a multi-chip module).

This chapter specifies an \emph{Advanced PLIC} that is not backward
compatible with the earlier {\RISCV} PLIC.
Full conformance to the Advanced Interrupt Architecture requires the
Advanced PLIC.
However, a workable system can be built substituting the older PLIC
instead, assuming only wired interrupts to harts, not MSIs.
Chapter~\ref{ch:DuoPLIC} describes a \emph{\mbox{Duo-PLIC\/}} that is
software-configurable to act as either an original {\RISCV} PLIC or an
Advanced PLIC.

\begin{commentary}
We intend eventually to provide a free example parameterized
implementation of an Advanced PLIC, written in portable SystemVerilog,
that we expect will be suitable for many {\RISCV} systems without
modification.
\end{commentary}

@@

An individual PLIC supports a fixed number of \emph{interrupt sources},
corresponding exactly with the set of physical incoming interrupt
wires at the PLIC.
Most often, each source's incoming wire is connected to the output
interrupt wire from a single device or device controller.
(For level-sensitive interrupts, the interrupt outputs of multiple
devices or controllers may be combined to drive the incoming wire of a
single interrupt source at a PLIC.
An interrupt source's incoming wire might also be simply tied high or
low, if, for example, the source will always be configured as Detached.
See Section~\ref{sec:AdvPLIC-reg-sourcecfg} for a description of
\emph{source modes}.)

Each of a PLIC's interrupt sources has a fixed unique
\emph{identity number} in the range 1 to~$N$, where $N$ is the total
number of sources at the PLIC.
The number zero is not a valid interrupt identity number at a PLIC.
The maximum number of interrupt sources an Advanced PLIC may support is
1023.

When a PLIC delivers interrupts directly to harts at a given
privilege level (rather than forwarding interrupts as MSIs), the PLIC
is the external interrupt controller for the harts at that privilege
level, and the interrupt identities at the PLIC become directly the
\emph{minor identities} for external interrupts at the harts.

On the other hand, when a PLIC forwards interrupts by MSIs, software
configures a new interrupt identity number for the outgoing MSIs of
each source.
Consequently, in this case, the source identity numbers at a given
PLIC only distinguish the incoming interrupts at the PLIC and have no
relevance outside the PLIC.

@@

An Advanced PLIC supports one or more \emph{interrupt domains}, each
associated with a subset of {\RISCV} harts at one privilege level
(machine or supervisor level).
The harts within an interrupt domain are those that the domain can
interrupt at the corresponding privilege level.
Each domain has its own memory-mapped control region in the machine's
address space that appears to control a complete, separate PLIC,
though in fact all domain interfaces together access a single combined
interrupt controller.

Figures \ref{fig:AdvPLIC-ex-1Domain} through
\ref{fig:AdvPLIC-ex-3Domains} depict some possible hierarchies of
interrupt domains implemented by an Advanced PLIC in a {\RISCV} system.

The first figure represents a minimal system that has a single hart not
supporting supervisor mode, with a single interrupt domain for machine
level on that hart.
The next figure, \ref{fig:AdvPLIC-ex-2Domains}, shows a basic
arrangement for a larger system designed for symmetric multiprocessing
(SMP), with multiple harts that all implement supervisor mode.
In such cases, the PLIC will usually provide a separate interrupt
domain for supervisor level, as the figure portrays.
This supervisor-level interrupt domain allows an operating system,
running in \mbox{S-mode} on the multiple harts, to have direct control
over the interrupts it receives, avoiding the need to call upon
\mbox{M-mode} to exercise that control.

@@

\begin{figure}[th]
\centerline{\includegraphics[scale=0.55]{AdvPLIC-ex-2Domains.png}}
\caption{%
An example system with four harts that implement \mbox{M-mode} and
\mbox{S-mode}, with two PLIC interrupt domains, one each for machine
and supervisor levels.%
}
\label{fig:AdvPLIC-ex-2Domains}
\end{figure}

A PLIC's interrupt domains are arranged in a tree hierarchy, with the
root domain always being at machine level.
Incoming interrupt wires arrive first at the root domain.
Each domain may then selectively delegate all or a subset of interrupt
sources to its child domains in the hierarchy.
Within a given PLIC, interrupt source numbers are invariant across all
domains, so source identity number~$i$ always refers to the same source
in every domain, corresponding to incoming wire number~$i$.
For an interrupt domain below the root, interrupt sources not delegated
down to that domain appear to the domain as being not implemented.

@@

\begin{commentary}
In order for the harts within an interrupt domain to have direct
control over the interrupts from the domain, the harts must be
cooperatively controlled by software at the same privilege level.
In particular, a single operating system should control all of the
harts associated with a supervisor-level interrupt domain.
In the examples of Figures \ref{fig:AdvPLIC-ex-2Domains}
and~\ref{fig:AdvPLIC-ex-3Domains}, control of the PLIC's
supervisor-level interrupt domain could not be safely split among
multiple independent OSes.

Given the domain hierarchies depicted in the figures, if it were
necessary to partition the application harts for multiple OSes,
machine-level software would need to prevent direct OS access to the
supervisor-level interrupt domain and instead provide SBI services
for controlling PLIC interrupts or, alternatively, emulate the control
interfaces of separate supervisor-level interrupt domains, one for each
OS.
Note that such emulation might still make use of the PLIC's
physical supervisor-level interrupt domain, but under the control of
machine-level software.
\end{commentary}

A PLIC's interrupt domain hierarchy satisfies these rules:
\begin{itemize}

@@

When a {\RISCV} hart's external interrupt controller is a PLIC, not an
IMSIC, the hart can be within only one interrupt domain of this PLIC at
each privilege level.

On the other hand, a hart that has an IMSIC for its external interrupt
controller may, at each privilege level, be in multiple PLIC interrupt
domains, even those of the same PLIC, and may potentially receive MSIs
from multiple different PLICs in the machine.

A platform might give software a way to choose between multiple
interrupt domain hierarchies for any given PLIC.
Any such configurability is outside the scope of this specification,
but should be available to machine level only.

@@

Within a given interrupt domain, each of the domain's harts has a
unique \emph{index number} in the range 0 to ${\mbox{2}^{14}-\mbox{1}}$
(=~16,383).
The index number a domain associates with a hart may or may not have
any relationship to the unique hart identifier (``hart ID'') that the
{\RISCV} Privileged Architecture assigns to the hart.
Two different interrupt domains may employ entirely different index
numbers for the same set of harts.
However, if any of a PLIC's interrupt domains can forward interrupts by
MSI, then all machine-level domains of the PLIC share a common mapping
of index numbers to harts.

@@

Each interrupt domain implemented by a PLIC has its own separate
physical control interface that is memory-mapped in the machine's
address space, allowing access to each domain to be easily regulated
by both PMP (physical memory protection) and page-based address
translation.
The control interfaces of all interrupt domains have a common
structure.
In most respects, every domain appears to software as though it were
a root domain, without visibility of the domains above it in the
hierarchy.

An individual interrupt domain has the following components for each
interrupt source at the PLIC:
\begin{itemize}

@@

\begin{commentary}
Although a PLIC with multiple interrupt domains may appear to duplicate
the per-source state listed above (source configuration, etc.) by a
factor equal to the number of domains, in fact, PLIC implementations
can exploit the fact that each source is ultimately active in only one
domain.
In all domains to which a specific interrupt source has not been
delegated, the state associated with the source appears as read-only
zeros, requiring no physical register bits.
\end{commentary}

@@

For each interrupt domain that a PLIC supports, there is a dedicated
memory-mapped control region for managing interrupts in that domain.
This control region is a multiple of 4~KiB in size and aligned to a
\mbox{4-KiB} address boundary.
The smallest valid control region is 16~KiB.
An interrupt domain's control region is populated by a set of
\mbox{32-bit} registers.
The first 16~KiB contains the registers listed in
Table~\ref{tab:AdvPLIC-domainControlRegion}.

@@

Starting at offset \z{0x4000}, an interrupt domain's control region
may optionally have an array of \emph{interrupt delivery control} (IDC)
structures, one for each potential hart index number in the range 0 to
some maximum that is at least as large as the maximum hart index number
for the interrupt domain.
IDC structures are used only when the domain is configured to deliver
interrupts directly to harts instead of being forwarded by MSIs.
An interrupt domain that supports only interrupt forwarding by MSIs
and not the direct delivery of interrupts by the PLIC does not need IDC
structures in its control region.

@@

The registers of the first 16~KiB of an interrupt domain's control
region (all but the IDC structures) are documented individually below.
IDC structures are documented later, in
Section~\ref{sec:AdvPLIC-directMode},
``Interrupt delivery directly by the PLIC.''

@@

Field DM (Delivery Mode) is {\WARL} and determines how this interrupt
domain delivers interrupts to harts.
The two possible values for DM are:
\begin{displayLinesTable}
0~=~direct delivery mode\\
1~=~MSI delivery mode\\
\end{displayLinesTable}
In \emph{direct delivery mode}, interrupts are prioritized and signaled
directly to harts by the PLIC itself.
In \emph{MSI delivery mode}, interrupts are forwarded by the PLIC
as MSIs to harts, presumably for further handling by IMSICs at those
harts.
A given PLIC implementation may support either or both of these
delivery modes for each interrupt domain.

@@

At system reset, all writable bits in \z{domaincfg} are initialized to
zero, including~IE.
If an implementation supports additional forms of reset for the PLIC,
it is implementation-defined (or possibly platform-defined) how these
other resets may affect \z{domaincfg}.

@@

If no interrupt domain of the PLIC supports MSI delivery
mode (\z{domaincfg}.DM is read-only zero for all domains),
these two registers are not implemented for any domain.
Otherwise, they are implemented for the root domain, and may
or may not be implemented for other machine-level domains.
For domains not at machine level, they are never implemented.
When a domain does not implement \z{mmsiaddrcfg} and
\z{mmsiaddrcfgh}, the eight bytes at their locations
are simply read-only zeros like other reserved bytes.

@@

If bit~L in \z{mmsiaddrcfgh} is set to one, \z{mmsiaddrcfg} and
\z{mmsiaddrcfgh} are \emph{locked}, and writes to the registers are
ignored, making the registers effectively read-only.
When L~=~1, the other fields in \z{mmsiaddrcfg} and \z{mmsiaddrcfgh}
may optionally all read as zeros.
In that case, if these other fields were given nonzero values
when L was first set in the root domain,
their values are retained internally by the
PLIC but become no longer visible by reading \z{mmsiaddrcfg} and
\z{mmsiaddrcfgh}.

@@

For the root domain, L~is initialized at
system reset to either zero or one, whichever is
deemed appropriate for the specific PLIC implementation.
If reset initializes L to one, either the other fields are
hardwired by the PLIC to constants, or the PLIC has a different means,
outside of this standard, for determining the addresses of outgoing MSI
writes.
In the latter case, the other fields in \z{mmsiaddrcfg} and
\z{mmsiaddrcfgh} may all read as zeros, so registers \z{mmsiaddrcfg} and
\z{mmsiaddrcfgh} have only read-only values zero and \z{0x80000000}
respectively.
Any time \z{mmsiaddrcfg} or \z{mmsiaddrcfgh} has a different value
(not zero or \z{0x80000000} respectively), the addresses for
outgoing MSI writes directed to machine level must be derivable
from the visible values of these registers, as specified in
Section~\ref{sec:AdvPLIC-MSIAddrs}.

@@

\begin{commentary}
Giving software the ability to arbitrarily determine the addresses to
which MSIs are sent, even if allowed only for machine level, permits
bypassing physical memory protection (PMP).
For PLICs that support MSI delivery mode, it is recommended, if
feasible, that the PLIC internally hardwire the physical addresses for
all target IMSICs, putting those addresses beyond the reach of software
to change.
However, not all PLIC implementations will be able to follow that
recommendation.

@@

PLICs that actually hardwire the IMSIC addresses internally
can implement these registers simply as read-only with values zero
and \z{0x80000000}.
Or, if the IMSIC addresses must be configured by software but
the formula is too complex for registers \z{mmsiaddrcfg} and
\z{mmsiaddrcfgh} to handle, again the registers can be implemented
simply as read-only with values zero and \z{0x80000000}, and a
separate, custom mechanism supplied for configuring the IMSIC
addresses.
\end{commentary}

If a PLIC supports additional forms of reset
besides system reset, it is implementation-defined (or possibly
platform-defined) how these other resets may affect \z{mmsiaddrcfg} and
\z{mmsiaddrcfgh} (as well as \z{smsiaddrcfg} and \z{smsiaddrcfgh})
in the root domain.
However, it must not be possible for insufficiently privileged software
to use a localized reset to unlock these registers by changing bit~L
back to zero.
For this reason, it is likely that only a complete system reset affects
these registers, and any other resets do not.

@@

Registers \z{smsiaddrcfg} and \z{smsiaddrcfgh} are implemented
by a domain if the domain implements \z{mmsiaddrcfg} and
\z{mmsiaddrcfgh} and the PLIC has at least one supervisor-level
interrupt domain.
If the registers are not implemented, the eight bytes at their
locations are simply read-only zeros like other reserved bytes.

@@

For the root domain only, if \z{mmsiaddrcfgh}.L =~1 and the
MSI-address-configuration fields are hidden (so \z{mmsiaddrcfgh}
reads as \z{0x80000000} and registers \z{mmsiaddrcfg},
\z{smsiaddrcfg}, and \z{smsiaddrcfgh} all read as zeros),
then whatever values \z{smsiaddrcfg} and \z{smsiaddrcfgh} had when
\z{mmsiaddrcfgh}.L was first set are retained internally by the PLIC,
though those values are no longer visible by reading the registers.
Alternatively, if system reset initializes \z{mmsiaddrcfgh}.L =~1 in
the root domain, and if all MSI-address-configuration fields never
appear as anything other than zeros, then the PLIC implementation
has some other, possibly nonstandard, means for determining
the addresses of outgoing MSIs, as discussed in the previous
subsection, \ref{sec:AdvPLIC-reg-mmsiaddrcfg}.

@@

When the interrupt domain is configured in MSI delivery mode
(\z{domaincfg}.DM =~1), register \z{genmsi} can be used to cause an
\emph{extempore} MSI to be sent from the PLIC to a hart.
The main purpose for this function is to assist in establishing
a temporary known ordering between a hart's writes to the PLIC's
registers and the transmission of MSIs from the PLIC to the hart, as
explained later in Section~\ref{sec:AdvPLIC-MSISync}.

\begin{commentary}
For other purposes, sending an MSI to a hart is usually better done by
writing directly to the hart's IMSIC, rather than employing a PLIC as
an intermediary.
Use of the \z{genmsi} register should be minimized to avoid it becoming
a bottleneck.
\end{commentary}

@@

A pending extempore MSI should be sent by the PLIC with minimal delay.
Once it has left the PLIC and the PLIC is able to accept a new write to
\z{genmsi} for another extempore MSI, Busy reverts to false.
All MSIs previously sent from this PLIC to the same hart must be
visible at the hart's IMSIC before the extempore MSI becomes visible at
the hart's IMSIC.

@@

Field IPRIO (Interrupt Priority) specifies the \emph{priority number}
for the interrupt source.
This field is a {\WARL} unsigned integer of \emph{IPRIOLEN} bits, where
IPRIOLEN is a constant parameter for the given PLIC, in the range of
1 to~8.
Only values 1 through $\mbox{2}^{\textrm{IPRIOLEN}} - \mbox{1}$ are
allowed for IPRIO, not zero.
A write to a \z{target} register sets IPRIO equal to bits
$({\mbox{IPRIOLEN} - \mbox{1}})$:0 of the \mbox{32-bit} value written,
unless those bits are all zeros, in which case the priority number is
set to~1 instead.
(If IPRIOLEN =~1, these rules cause IPRIO to be effectively read-only
with value~1.)

@@

\item
The pending bit is cleared when the interrupt is claimed at the PLIC or
forwarded by MSI, or by a relevant write to an \z{in\_clrip} register
or to \z{clripnum}.

@@

\item
The pending bit is cleared whenever the rectified input value is low.
The pending bit is not cleared by a claim of the interrupt at the PLIC,
nor can it be cleared by a write to an \z{in\_clrip} register or to
\z{clripnum}.

@@

%-----------------------------------------------------------------------
\section{Interrupt delivery directly by the PLIC}
\label{sec:AdvPLIC-directMode}

>>>>

%=======================================================================
\chapter{Advanced Platform-Level Interrupt Controller (APLIC)}
\label{ch:AdvPLIC}
\chaptermark{APLIC}

@@

This chapter specifies an \emph{Advanced PLIC} (APLIC) that is not backward
compatible with the earlier {\RISCV} PLIC.
Full conformance to the Advanced Interrupt Architecture requires the
APLIC.
However, a workable system can be built substituting the older PLIC
instead, assuming only wired interrupts to harts, not MSIs.
Chapter~\ref{ch:DuoPLIC} describes a \emph{\mbox{Duo-PLIC\/}} that is
software-configurable to act as either an original {\RISCV} PLIC or an
APLIC.

\begin{commentary}
We intend eventually to provide a free example parameterized
implementation of an APLIC, written in portable SystemVerilog,
that we expect will be suitable for many {\RISCV} systems without
modification.
\end{commentary}

In a machine without IMSICs, every {\RISCV} hart accepts interrupts
from exactly one PLIC or APLIC that is the \emph{external interrupt controller}
for that hart.
A hart's external interrupt controller (the PLIC or APLIC) signals interrupts
to the hart through a dedicated connection, usually a wire, for each
privilege level that the hart may receive interrupts.
(Recall Figure~\ref{fig:intrsWithoutIMSICs} on
page~\pageref{fig:intrsWithoutIMSICs}.)
A system without IMSICs will typically have only one PLIC or APLIC, serving as
the external interrupt controller for all {\RISCV} harts.

\begin{commentary}
Because every {\RISCV} hart without an IMSIC has exactly one PLIC or APLIC
as its external interrupt controller, a system with multiple APLICs
must partition the harts into disjoint subsets, making each APLIC the
external interrupt controller for a separate subset of the harts.
While not prohibited, this arrangement is likely to be less efficient
than having all harts share a single APLIC.
\end{commentary}

{\RISCV} harts that employ IMSICs as their external interrupt
controllers can receive external interrupts only in the form of MSIs.
In that case, the role of an APLIC is to convert wired interrupts into
MSIs for harts.
(Recall Figure~\ref{fig:intrsWithIMSICs} on
page~\pageref{fig:intrsWithIMSICs}.)
The APLIC is said to \emph{forward} incoming wire-signaled interrupts to
harts by sending MSIs to the harts.

When harts have IMSICs to support MSIs, a system may easily contain
multiple APLICs for converting wired interrupts into MSIs, with each
APLIC forwarding interrupts from a different subset of devices.
Multiple APLICs are presumably more likely to arise when groups of
devices are physically distant from one another, perhaps even on
separate chips (including chiplets in a multi-chip module).

@@

An individual APLIC supports a fixed number of \emph{interrupt sources},
corresponding exactly with the set of physical incoming interrupt
wires at the APLIC.
Most often, each source's incoming wire is connected to the output
interrupt wire from a single device or device controller.
(For level-sensitive interrupts, the interrupt outputs of multiple
devices or controllers may be combined to drive the incoming wire of a
single interrupt source at an APLIC.
An interrupt source's incoming wire might also be simply tied high or
low, if, for example, the source will always be configured as Detached.
See Section~\ref{sec:AdvPLIC-reg-sourcecfg} for a description of
\emph{source modes}.)

Each of an APLIC's interrupt sources has a fixed unique
\emph{identity number} in the range 1 to~$N$, where $N$ is the total
number of sources at the APLIC.
The number zero is not a valid interrupt identity number at an APLIC.
The maximum number of interrupt sources an APLIC may support is
1023.

When an APLIC delivers interrupts directly to harts at a given
privilege level (rather than forwarding interrupts as MSIs), the APLIC
is the external interrupt controller for the harts at that privilege
level, and the interrupt identities at the APLIC become directly the
\emph{minor identities} for external interrupts at the harts.

On the other hand, when an APLIC forwards interrupts by MSIs, software
configures a new interrupt identity number for the outgoing MSIs of
each source.
Consequently, in this case, the source identity numbers at a given
APLIC only distinguish the incoming interrupts at the APLIC and have no
relevance outside the APLIC.

@@

An APLIC supports one or more \emph{interrupt domains}, each
associated with a subset of {\RISCV} harts at one privilege level
(machine or supervisor level).
The harts within an interrupt domain are those that the domain can
interrupt at the corresponding privilege level.
Each domain has its own memory-mapped control region in the machine's
address space that appears to control a complete, separate APLIC,
though in fact all domain interfaces together access a single combined
interrupt controller.

Figures \ref{fig:AdvPLIC-ex-1Domain} through
\ref{fig:AdvPLIC-ex-3Domains} depict some possible hierarchies of
interrupt domains implemented by an APLIC in a {\RISCV} system.

The first figure represents a minimal system that has a single hart not
supporting supervisor mode, with a single interrupt domain for machine
level on that hart.
The next figure, \ref{fig:AdvPLIC-ex-2Domains}, shows a basic
arrangement for a larger system designed for symmetric multiprocessing
(SMP), with multiple harts that all implement supervisor mode.
In such cases, the APLIC will usually provide a separate interrupt
domain for supervisor level, as the figure portrays.
This supervisor-level interrupt domain allows an operating system,
running in \mbox{S-mode} on the multiple harts, to have direct control
over the interrupts it receives, avoiding the need to call upon
\mbox{M-mode} to exercise that control.

@@

\begin{figure}[th]
\centerline{\includegraphics[scale=0.55]{AdvPLIC-ex-2Domains.png}}
\caption{%
An example system with four harts that implement \mbox{M-mode} and
\mbox{S-mode}, with two APLIC interrupt domains, one each for machine
and supervisor levels.%
}
\label{fig:AdvPLIC-ex-2Domains}
\end{figure}

An APLIC's interrupt domains are arranged in a tree hierarchy, with the
root domain always being at machine level.
Incoming interrupt wires arrive first at the root domain.
Each domain may then selectively delegate all or a subset of interrupt
sources to its child domains in the hierarchy.
Within a given APLIC, interrupt source numbers are invariant across all
domains, so source identity number~$i$ always refers to the same source
in every domain, corresponding to incoming wire number~$i$.
For an interrupt domain below the root, interrupt sources not delegated
down to that domain appear to the domain as being not implemented.

@@

\begin{commentary}
In order for the harts within an interrupt domain to have direct
control over the interrupts from the domain, the harts must be
cooperatively controlled by software at the same privilege level.
In particular, a single operating system should control all of the
harts associated with a supervisor-level interrupt domain.
In the examples of Figures \ref{fig:AdvPLIC-ex-2Domains}
and~\ref{fig:AdvPLIC-ex-3Domains}, control of the APLIC's
supervisor-level interrupt domain could not be safely split among
multiple independent OSes.

Given the domain hierarchies depicted in the figures, if it were
necessary to partition the application harts for multiple OSes,
machine-level software would need to prevent direct OS access to the
supervisor-level interrupt domain and instead provide SBI services
for controlling APLIC interrupts or, alternatively, emulate the control
interfaces of separate supervisor-level interrupt domains, one for each
OS.
Note that such emulation might still make use of the APLIC's
physical supervisor-level interrupt domain, but under the control of
machine-level software.
\end{commentary}

An APLIC's interrupt domain hierarchy satisfies these rules:
\begin{itemize}

@@

When a {\RISCV} hart's external interrupt controller is an APLIC, not an
IMSIC, the hart can be within only one interrupt domain of this APLIC at
each privilege level.

On the other hand, a hart that has an IMSIC for its external interrupt
controller may, at each privilege level, be in multiple APLIC interrupt
domains, even those of the same APLIC, and may potentially receive MSIs
from multiple different APLICs in the machine.

A platform might give software a way to choose between multiple
interrupt domain hierarchies for any given APLIC.
Any such configurability is outside the scope of this specification,
but should be available to machine level only.

@@

Within a given interrupt domain, each of the domain's harts has a
unique \emph{index number} in the range 0 to ${\mbox{2}^{14}-\mbox{1}}$
(=~16,383).
The index number a domain associates with a hart may or may not have
any relationship to the unique hart identifier (``hart ID'') that the
{\RISCV} Privileged Architecture assigns to the hart.
Two different interrupt domains may employ entirely different index
numbers for the same set of harts.
However, if any of an APLIC's interrupt domains can forward interrupts by
MSI, then all machine-level domains of the APLIC share a common mapping
of index numbers to harts.

@@

Each interrupt domain implemented by an APLIC has its own separate
physical control interface that is memory-mapped in the machine's
address space, allowing access to each domain to be easily regulated
by both PMP (physical memory protection) and page-based address
translation.
The control interfaces of all interrupt domains have a common
structure.
In most respects, every domain appears to software as though it were
a root domain, without visibility of the domains above it in the
hierarchy.

An individual interrupt domain has the following components for each
interrupt source at the APLIC:
\begin{itemize}

@@

\begin{commentary}
Although an APLIC with multiple interrupt domains may appear to duplicate
the per-source state listed above (source configuration, etc.) by a
factor equal to the number of domains, in fact, APLIC implementations
can exploit the fact that each source is ultimately active in only one
domain.
In all domains to which a specific interrupt source has not been
delegated, the state associated with the source appears as read-only
zeros, requiring no physical register bits.
\end{commentary}

@@

For each interrupt domain that an APLIC supports, there is a dedicated
memory-mapped control region for managing interrupts in that domain.
This control region is a multiple of 4~KiB in size and aligned to a
\mbox{4-KiB} address boundary.
The smallest valid control region is 16~KiB.
An interrupt domain's control region is populated by a set of
\mbox{32-bit} registers.
The first 16~KiB contains the registers listed in
Table~\ref{tab:AdvPLIC-domainControlRegion}.

@@

Starting at offset \z{0x4000}, an interrupt domain's control region
may optionally have an array of \emph{interrupt delivery control} (IDC)
structures, one for each potential hart index number in the range 0 to
some maximum that is at least as large as the maximum hart index number
for the interrupt domain.
IDC structures are used only when the domain is configured to deliver
interrupts directly to harts instead of being forwarded by MSIs.
An interrupt domain that supports only interrupt forwarding by MSIs
and not the direct delivery of interrupts by the APLIC does not need IDC
structures in its control region.

@@

The registers of the first 16~KiB of an interrupt domain's control
region (all but the IDC structures) are documented individually below.
IDC structures are documented later, in
Section~\ref{sec:AdvPLIC-directMode},
``Interrupt delivery directly by the APLIC.''

@@

Field DM (Delivery Mode) is {\WARL} and determines how this interrupt
domain delivers interrupts to harts.
The two possible values for DM are:
\begin{displayLinesTable}
0~=~direct delivery mode\\
1~=~MSI delivery mode\\
\end{displayLinesTable}
In \emph{direct delivery mode}, interrupts are prioritized and signaled
directly to harts by the APLIC itself.
In \emph{MSI delivery mode}, interrupts are forwarded by the APLIC
as MSIs to harts, presumably for further handling by IMSICs at those
harts.
A given APLIC implementation may support either or both of these
delivery modes for each interrupt domain.

@@

At system reset, all writable bits in \z{domaincfg} are initialized to
zero, including~IE.
If an implementation supports additional forms of reset for the APLIC,
it is implementation-defined (or possibly platform-defined) how these
other resets may affect \z{domaincfg}.

@@

If no interrupt domain of the APLIC supports MSI delivery
mode (\z{domaincfg}.DM is read-only zero for all domains),
these two registers are not implemented for any domain.
Otherwise, they are implemented for the root domain, and may
or may not be implemented for other machine-level domains.
For domains not at machine level, they are never implemented.
When a domain does not implement \z{mmsiaddrcfg} and
\z{mmsiaddrcfgh}, the eight bytes at their locations
are simply read-only zeros like other reserved bytes.

@@

If bit~L in \z{mmsiaddrcfgh} is set to one, \z{mmsiaddrcfg} and
\z{mmsiaddrcfgh} are \emph{locked}, and writes to the registers are
ignored, making the registers effectively read-only.
When L~=~1, the other fields in \z{mmsiaddrcfg} and \z{mmsiaddrcfgh}
may optionally all read as zeros.
In that case, if these other fields were given nonzero values
when L was first set in the root domain,
their values are retained internally by the
APLIC but become no longer visible by reading \z{mmsiaddrcfg} and
\z{mmsiaddrcfgh}.

@@

For the root domain, L~is initialized at
system reset to either zero or one, whichever is
deemed appropriate for the specific APLIC implementation.
If reset initializes L to one, either the other fields are
hardwired by the APLIC to constants, or the APLIC has a different means,
outside of this standard, for determining the addresses of outgoing MSI
writes.
In the latter case, the other fields in \z{mmsiaddrcfg} and
\z{mmsiaddrcfgh} may all read as zeros, so registers \z{mmsiaddrcfg} and
\z{mmsiaddrcfgh} have only read-only values zero and \z{0x80000000}
respectively.
Any time \z{mmsiaddrcfg} or \z{mmsiaddrcfgh} has a different value
(not zero or \z{0x80000000} respectively), the addresses for
outgoing MSI writes directed to machine level must be derivable
from the visible values of these registers, as specified in
Section~\ref{sec:AdvPLIC-MSIAddrs}.

@@

\begin{commentary}
Giving software the ability to arbitrarily determine the addresses to
which MSIs are sent, even if allowed only for machine level, permits
bypassing physical memory protection (PMP).
For APLICs that support MSI delivery mode, it is recommended, if
feasible, that the APLIC internally hardwire the physical addresses for
all target IMSICs, putting those addresses beyond the reach of software
to change.
However, not all APLIC implementations will be able to follow that
recommendation.

@@

APLICs that actually hardwire the IMSIC addresses internally
can implement these registers simply as read-only with values zero
and \z{0x80000000}.
Or, if the IMSIC addresses must be configured by software but
the formula is too complex for registers \z{mmsiaddrcfg} and
\z{mmsiaddrcfgh} to handle, again the registers can be implemented
simply as read-only with values zero and \z{0x80000000}, and a
separate, custom mechanism supplied for configuring the IMSIC
addresses.
\end{commentary}

If an APLIC supports additional forms of reset
besides system reset, it is implementation-defined (or possibly
platform-defined) how these other resets may affect \z{mmsiaddrcfg} and
\z{mmsiaddrcfgh} (as well as \z{smsiaddrcfg} and \z{smsiaddrcfgh})
in the root domain.
However, it must not be possible for insufficiently privileged software
to use a localized reset to unlock these registers by changing bit~L
back to zero.
For this reason, it is likely that only a complete system reset affects
these registers, and any other resets do not.

@@

Registers \z{smsiaddrcfg} and \z{smsiaddrcfgh} are implemented
by a domain if the domain implements \z{mmsiaddrcfg} and
\z{mmsiaddrcfgh} and the APLIC has at least one supervisor-level
interrupt domain.
If the registers are not implemented, the eight bytes at their
locations are simply read-only zeros like other reserved bytes.

@@

For the root domain only, if \z{mmsiaddrcfgh}.L =~1 and the
MSI-address-configuration fields are hidden (so \z{mmsiaddrcfgh}
reads as \z{0x80000000} and registers \z{mmsiaddrcfg},
\z{smsiaddrcfg}, and \z{smsiaddrcfgh} all read as zeros),
then whatever values \z{smsiaddrcfg} and \z{smsiaddrcfgh} had when
\z{mmsiaddrcfgh}.L was first set are retained internally by the APLIC,
though those values are no longer visible by reading the registers.
Alternatively, if system reset initializes \z{mmsiaddrcfgh}.L =~1 in
the root domain, and if all MSI-address-configuration fields never
appear as anything other than zeros, then the APLIC implementation
has some other, possibly nonstandard, means for determining
the addresses of outgoing MSIs, as discussed in the previous
subsection, \ref{sec:AdvPLIC-reg-mmsiaddrcfg}.

@@

When the interrupt domain is configured in MSI delivery mode
(\z{domaincfg}.DM =~1), register \z{genmsi} can be used to cause an
\emph{extempore} MSI to be sent from the APLIC to a hart.
The main purpose for this function is to assist in establishing
a temporary known ordering between a hart's writes to the APLIC's
registers and the transmission of MSIs from the APLIC to the hart, as
explained later in Section~\ref{sec:AdvPLIC-MSISync}.

\begin{commentary}
For other purposes, sending an MSI to a hart is usually better done by
writing directly to the hart's IMSIC, rather than employing an APLIC as
an intermediary.
Use of the \z{genmsi} register should be minimized to avoid it becoming
a bottleneck.
\end{commentary}

@@

A pending extempore MSI should be sent by the APLIC with minimal delay.
Once it has left the APLIC and the APLIC is able to accept a new write to
\z{genmsi} for another extempore MSI, Busy reverts to false.
All MSIs previously sent from this APLIC to the same hart must be
visible at the hart's IMSIC before the extempore MSI becomes visible at
the hart's IMSIC.

@@

Field IPRIO (Interrupt Priority) specifies the \emph{priority number}
for the interrupt source.
This field is a {\WARL} unsigned integer of \emph{IPRIOLEN} bits, where
IPRIOLEN is a constant parameter for the given APLIC, in the range of
1 to~8.
Only values 1 through $\mbox{2}^{\textrm{IPRIOLEN}} - \mbox{1}$ are
allowed for IPRIO, not zero.
A write to a \z{target} register sets IPRIO equal to bits
$({\mbox{IPRIOLEN} - \mbox{1}})$:0 of the \mbox{32-bit} value written,
unless those bits are all zeros, in which case the priority number is
set to~1 instead.
(If IPRIOLEN =~1, these rules cause IPRIO to be effectively read-only
with value~1.)

@@

\item
The pending bit is cleared when the interrupt is claimed at the APLIC or
forwarded by MSI, or by a relevant write to an \z{in\_clrip} register
or to \z{clripnum}.

@@

\item
The pending bit is cleared whenever the rectified input value is low.
The pending bit is not cleared by a claim of the interrupt at the APLIC,
nor can it be cleared by a write to an \z{in\_clrip} register or to
\z{clripnum}.

@@

%-----------------------------------------------------------------------
\section{Interrupt delivery directly by the APLIC}
\label{sec:AdvPLIC-directMode}

When an interrupt domain is in direct delivery mode
(\z{domaincfg}.DM =~0), interrupts are delivered from the APLIC to harts
by a unique signal to each hart, usually a dedicated wire.
In this case, the domain's memory-mapped control region contains at the
end an array of interrupt delivery control (IDC) structures, one IDC
structure per potential hart index.
The first IDC structure is for the domain's hart with index~0;
the second is for the hart with index~1; etc.
